---
title: "Final Project: State/Region Incarceration Levels by Sex"
author: "Ava, Alex, Luci, and Christine"
date: "6-7-26"
output: html_document
---

## Data Processing, Methods, and Figures

### Data Processing

Our class readings focused on both the practical ways to code data visualization, but also on the methods through which we choose to share data. In sociology, data visualizations have been historically important. For example, Du Bois’ work shows how he depicted and interrogated racial inequality through data visualizations and utilized spatial data to identify water pumps associated with cholera (Battle-Baptiste and Rusert 2018). Our group took these class lessons and readings into consideration when discussing how we should show the data.
	
For our methods, we employed several tools in R to analyze the data. We used the Bureau of Justice (BJS) statistics data from 2023, which measured the number of people incarcerated in each state in the United States, and we loaded this dataset into our R markdown file. Our BJS data is “big” and it was also “dirty,” which means that it includes information that is not relevant or applicable to our research question (Salganick 2017). This means that we were required to pre-process and “clean” the data before we could apply our code and create our data visualizations. We used R’s “mutate” and “case_match” feature to do this. From the dataset, there were several state names that needed to be edited because they had “/b” attached to the state’s name (e.g., “Alaska/b” needed to become “Alaska”) through the mutate feature to match tidycensus data and for presentation purposes. The data signifiers, although providing valuable information on differences about how the number of incarcerated individuals was counted, needed to be edited to pair the data through tidycensus and for the overall aesthetics. We address these differences on how the number of incarcerated individuals was counted in our limitations. Moreover, we also needed to get rid of the “U.S. total,” “Federal,” and “All States” incarceration rates from the data, as we were only looking at incarceration rates from individual states. In addition, we dropped Washington D.C. and Puerto Rico from the dataset because they had no available data on those locations and they are not states. This dataset is titled “data” in our R markdown file. 

We also wanted to look at the data by population size in each state and region because we were interested to see how different policies, cultures, and locations can influence incarceration rates. In order to do this and ensure that we were analyzing the data proportionally to the population, we had to create a measure on how many females, males, and individuals were being incarcerated per 100,000 people in each state. This measure is important because only looking at the total number of people incarcerated in the state might be misleading, as some states’ higher population size might lead to higher numbers of people incarcerated. Thus, we sought to find the proportion of people incarcerated compared to the total population in the state. In order to do this we used tidycensus and an API Key to get population data for the year 2023 using the variable “B01003_001”, the same year as our incarceration data. We filtered out Washington D.C. and Puerto Rico here as well as we did not need them in the dataset. Then, we joined the datasets using left_join() using the column titles for each dataset (NAME for tidycensus and Location for BJS).
	
Download all necessary libraries
```{r libraries, echo = T, message=F, warning=F}
#Libraries
library(RCurl)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)
library(tidycensus)
library(tigris)
library(viridis)
library(magrittr)

```


Download data 
```{r data, echo = T, message=F, warning=F}

url <- getURL("https://raw.githubusercontent.com/avaelisa/soc10finalproject-alex-christine-ava-luci/refs/heads/main/soc10-dataset")

# download incarceration data from the BJS (from github)
data <- read_csv(url)

head(data) #uncleaned data downloaded from the BJS

```

Data processing part 1: fix names, prep per 100,000 people, and spatial data 

```{r data processing 1, echo = T, message=F, warning=F}

#Mutate to fix state names (remove data type signifiers) and remove U.S. total, Federal, and All 
#States from BJS dataset
data %<>%
  mutate(Location=case_match(Location,
                             "Alaska/b"~"Alaska",
                             "Connecticut/b"~"Connecticut",
                             "Delaware/b"~"Delaware",
                             "Hawaii/b"~"Hawaii",
                             "Maryland/c"~"Maryland",
                             "Missouri/d"~"Missouri",
                             "Oklahoma/c"~"Oklahoma",
                             "Rhode Island/b"~"Rhode Island",
                             "Vermont/b"~"Vermont",
                             .default=Location)) %>%
                            drop_na(Total) %>% #Get rid of Washington D.C. and Puerto Rico 
  filter(!Location %in% c("U.S. total", "Federal", "All States"))

head(data) #Cleaned data
```

```{r data processing 1A, echo = T, message=F, warning=F, results= 'hide'}

#Census API
census_api_key("857ead205d4b6101ae723a246e9b6f95b1b827d4")

#Add spatial data and estimate data with tidycensus
states <- get_acs(
  geography = "state",
  variables = "B01003_001",
  geometry = TRUE,
  year = 2023) %>%
  filter(!NAME %in% c("District of Columbia", "Puerto Rico")) #remove these here as well
```

```{r data processing 1B, echo = T, message=F, warning=F}

head(states) #Tidycensus state data from 2023 (matching 2023 BJS data)

#Combine data with left_join USA map + incarceration rate
states_incar <- states %>%
  left_join(data, by = c("NAME" = "Location"))

head(states_incar) #Combined tidycensus and BJS data 

```

After combining the datasets, we wanted to use this information to create new columns/variabels for comparison. We then created a new row where we divided the number of females incarcerated per state by the estimate (or population) in the state, and multiplied this by 100,000. This is under the dataset “states_incar_fmpc” (number of females and males incarcerated per one hundred thousand people). After, we added a column for the total number of men incarcerated per 100,000 people in each state by doing the same thing but using the total number of males incarcerated in each state as opposed to the total number of females (“states_incar_fmpc”).  We then added a column for the total number of people incarcerated per 100,000 people in each state by doing the same thing but using the total number of people incarcerated in each state as opposed to the total number of females (“states_incar_ipp”). Each time, we used the previous dataset to create the following dataset with the new column.

```{r data processing 2, echo = T, message=F, warning=F}

#add number of females incarcerated per one hundred thousand people to the spatial data
states_incar_fpc <- states_incar %>%
  mutate("Number of Females Incarcerated Per 100,000 People" = Female/estimate*100000)

head(states_incar_fpc) #New data set includes Number of Females Incarcerated Per 100,000 
#People in Each State in state population

#Create column of number of males incarcerated per 1000,000 people into dataset with 
#spatial data and Number of Females Incarcerated Per 100,000 People in Each State
states_incar_fmpc <- states_incar_fpc %>%
  mutate("Number of Males Incarcerated Per 100,000 People" = Male/estimate*100000)

head(states_incar_fmpc) #New data set includes number males incarcerated per 100,000 
#people in the state population

#total number of people incarcerated per population in each state
states_incar_ipp <- states_incar_fmpc %>%
  mutate("Number of People Incarcerated Per 100,000 People" = Total/estimate*100000)
```

We also focused on region data per 100,000 people. We pre-processed the data by finding the number of females and males who are incarcerated in each region per 100,000 people in each region. The states in each region can be found in the code below. Note that since Hawaii and Alaska are not part of any of the regions, we categorized them under “Other”. We used mutate() t to group the regions and then used group_by() and sum() to group the regions and the sum of the total number of people incarcerated (incarsum), total population (popsum), total number of females incarcerated (fincarsum), and total number of males incarcerated (mincarsum) in each region. We created dataset “states_incar_region” with this grouping and variable creation. Then, we used this data set to create “states_incar_region2”, in which we would use mutate() to get the proportional data of incarceration in each region. The calculation for this was dividing each of the sums for each group (male, female, total) by the total population times 100,000. This gave us the total number of people in each region, giving us the ratio for the number of females/males/total population incarcerated per 100,000 people in each region. 

```{r data processing 3, echo = T, message=F, warning=F}

#preparing region data 
states_incar_region <- states_incar_ipp %>% 
  mutate(Region=case_match(NAME,
                             #Northeast
                             c("Maine", "Vermont", "New Hampshire", "Massachusetts", "Connecticut", 
                               "Rhode Island", "New York", "Pennsylvania", "New Jersey") 
                           ~ "Northeast",
                             #Midwest
                             c("North Dakota", "South Dakota", "Nebraska", "Kansas", "Minnesota", 
                               "Iowa", "Missouri", "Illinois", "Wisconsin", "Michigan", "Ohio", 
                               "Indiana") ~ "Midwest",
                             #West
                             c("Washington", "Oregon", "California", "Nevada", "Idaho", "Wyoming", 
                               "Montana", "Colorado", "Utah", "Arizona", "New Mexico") ~ "West",
                             #South
                             c("Texas", "Oklahoma", "Arkansas", "Louisiana", "Mississippi", 
                               "Kentucky", "Tennessee", "Alabama", "Georgia", "West Virginia", 
                               "Florida", "South Carolina", "North Carolina", "Maryland", 
                               "Delaware","Virginia") ~ "South",
                             .default = "Other")) %>% #alaska and hawaii
                            drop_na(Total) %>%
  group_by(Region) %>% mutate(incarsum = sum(Total, na.rm = TRUE)) %>%
  group_by(Region) %>% mutate(popsum = sum(estimate, na.rm = TRUE)) %>%
  group_by(Region) %>% mutate(fincarsum = sum(Female, na.rm = TRUE)) %>%
  group_by(Region) %>% mutate(mincarsum = sum(Male, na.rm = TRUE)) 

states_incar_region2 <- states_incar_region %>%
 mutate("Number of Females Incarcerated Per 100,000 People" = fincarsum/popsum*100000)%>%
 mutate("Number of Males Incarcerated Per 100,000 People" = mincarsum/popsum*100000) %>%
 mutate("Number of People Incarcerated Per 100,000 People" = incarsum/popsum*100000)
  
head(states_incar_region2) #States' data are now organized into their respective regions 
#with the calculations of each of the sums in a column according to their region

```

### Methods and Figures

After pre-processing the data, we were able to create visual bar plots. The first bar chart (Bar Chart 1A) we created compared the incarceration rates for all states. We took the preprocessed data and used ggplot to create the bar charts, which shows each state’s total incarceration rate. Each state is color-coded with a different color, shown through the legend. We used “coord_flip” to showcase the data horizontally because we believed this showed the data more clearly than vertically. 

### *Bar Charts 1A, 1B, and 1C*

Bar Chart 1A: Total Population Incarcerated by State
Goal: Compare incarceration totals across all states

Using coord_flip to make bar chart horizontal. Reorder in ggplot will rank bars from smallest to biggest so it is easier to see trends and which states have highest vs lowest incarceration rates. This will make data easier to analyze. 

```{r figure 1A, echo = T, message=F, warning=F}

state_total_graph <- ggplot(states_incar_ipp, aes(x = reorder(NAME, Total), y = Total))+ #Reorder 
  geom_bar(position = "dodge", stat = "identity")+
           labs(y = "Total", x="State", title = "Total Population Incarcerated by State")+
             coord_flip() +
  theme(legend.position = "none")


#See visualization 
state_total_graph 

```

To create Bar Chart 1B, we did the same thing as Bar Chart 1A, but utilized the variable “Number of People Incarcerated Per 100,000 People” instead of “Total”, as this would show us the number of people incarcerated in each state in relation to its total population. 

Bar Chart 1B: Number of People Incarcerated Per 100,000 People in the State
Goal: Compare total incarceration according to population for all states

```{r figure 1b, echo = T, message=F, warning=F }
        
state_pop_ratio_graph <- ggplot(states_incar_ipp, 
                                aes(x = reorder(NAME, 
                                `Number of People Incarcerated Per 100,000 People`), 
                               y = `Number of People Incarcerated Per 100,000 People`))+ #reorder 
  geom_bar(position = "dodge", stat = "identity")+
           labs(x = "State", y = "Total", 
                title = "Number of People Incarcerated/100,000 People in Each State")+
             coord_flip() +
  theme(legend.position = "none")


#See visualization 
state_pop_ratio_graph


```

Bar Chart 1C begins to look into the female incarceration data and the total population of a state. Using pre-processed data “states_incar_ipp”, we plotted the number of females incarcerated per 100,000 people in the state’s total population. Through ggplot, this data was displayed through a bar graph for each state. 

Bar Chart 1C: Number of Females Incarcerated Per 100,000 People in the State
Goal: Compare female incarceration according to population for all states

```{r figure 1C, echo = T, message=F, warning=F}

state_female_ratio_graph <- ggplot(states_incar_ipp, 
    aes(x = reorder(NAME, `Number of Females Incarcerated Per 100,000 People`), 
        y = `Number of Females Incarcerated Per 100,000 People`))+ #reorder 
  geom_bar(position = "dodge", stat = "identity") +
           labs(x = "State", y = "Total", 
                title = "Number of Females Incarcerated Per 100,000 People 
in Each State")+
             coord_flip() +
  theme(legend.position = "none")


#See visualization 
state_female_ratio_graph

```

Overall, the bar charts from section 1 show the total number of people incarcerated as well as how many total people and females are incarcerated per 100,000 people within each state. The significance of this data visualization is that although mass incarceration is widely considered a federal trend, this data indicates that some states have much lower incarceration numbers than others. This data analysis points to how a few states may be disproportionately contributing to the total federal incarceration numbers. 

### *Bar Charts 2A, 2B, and 2C*

For Bar Chart 2A, used our precleaned dataset called “states_incar_region2” to organize create a bar chart of the total number of people incarcerated (“incarsum”) in different regions of the United States: “West,” “South,” “Midwest,” and “Northeast.” Hawaii and Alaska are counted together as “Other”. By adding up the number of people incarcerated in the states for each region, these functions then provided us the total number of people who were incarcerated in each region. Each region is color-coded, and there are labels that identify which region is which.  

Bar Chart 2A: Total Number of People Incarcerated for Each Region (and preparing region data)
Goal: Divide state data into their respective regions (Northeast, Midwest, West, South) and add number of people incarcerated for each region

The states were divided into regions through case_match. Each state's incarceration numbers for each region were summed to calculate total incarceration numbers for each region.  

```{r figure 2, echo = T, message=F, warning=F}

#Plot
region_total_graph <- 
  ggplot(states_incar_region2, aes(x = reorder(Region, incarsum), y = incarsum)) +
  geom_bar(aes(fill = Region), position = "dodge", stat = "identity")+
  labs(x = "Region", y = "Total", title = "Total Number of People Incarcerated by Region")+
  theme_linedraw()+
  coord_flip()

#See visualization
region_total_graph

```

For Barchart 2B, we again used our precleaned dataset “states_incar_region2”. We completed the same steps as bar chart 2A; however, this time we used the variable “Number of People Incarcerated Per 100,000 People”. This allowed us to create a bar chart on the number of people incarcerated per 100,000 people in each region,

Bar Chart 2B: Number of People Incarcerated Per 100,000 People in Population for Each Region 
Goal: To show the Number of people incarcerated per 100,000 people in each region. This is important because this data takes into consideration total population size.  

```{r figure 2B, echo = T, message=F, warning=F}

#Plot
region_ipp_graph <- 
  ggplot(states_incar_region2, 
         aes(x = reorder(Region, `Number of People Incarcerated Per 100,000 People`), 
             y = `Number of People Incarcerated Per 100,000 People`)) +
  geom_bar(aes(fill = Region), position = "dodge", stat = "identity")+
  labs(x = "Region", y = "Number of People Incarcerated Per 100,000 People in Each Region", 
       title = "Number of People Incarcerated Per 100,000 People in Each Region")+
  theme_linedraw()+
  coord_flip()

#Show visualization
region_ipp_graph

```

For Barchart 2C, we measured the number of females incarcerated per 100,000 people in each region. We used the same code as Bar Chart 2A and 2B, except we changed the variable displayed to be the sum of the total number of females incarcerated per 100,000 people instead of total/100,000 people. This provided another bar chart highlighting the differences in the number of females each state incarcerates. 

Bar Chart 2C: Number of Females Incarcerated Per 100,000 People in Population for Each Region 
Goal: To show the Number of females incarcerated per 100,000 people in each region. This is important because this data takes into consideration total population size.  

```{r figure 2C, echo = T, message=F, warning=F}

#Plot
region_totals_graph <- 
  ggplot(states_incar_region2, 
         aes(x = reorder(Region, `Number of Females Incarcerated Per 100,000 People`), 
             y = `Number of Females Incarcerated Per 100,000 People`)) +
  geom_bar(aes(fill = Region), position = "dodge", stat = "identity")+
  labs(x = "Region", y = "Number of Females Incarcerated Per 100,000 People in Each Region", 
       title = "Number of Females Incarcerated Per 100,000 People in Each Region")+
  theme_linedraw()+
  coord_flip()

#Show visualization
region_totals_graph

```

Overall, the bar charts from section 2 identify the total number of people incarcerated as well as the number of people/females incarcerated per 100,000 people within each region. It is important to view the data visualizations that can show trends between proximate states and their total numbers of incarceration. Given that some regions have disproportionately higher incarceration numbers than others, this may point to regional-based differences that can explain disparate incarceration numbers. 

### *Stacked Bar Charts 3A and 3B*

For Barchart 3A, we provide a stacked bar chart to show the number of females incarcerated alongside the number of males incarcerated in each state. First, the variables in our state data were “pivoted longer” through the R function “pivot_longer” and the forward pipe operator to make the columns “Male” and “Female” from the “states_incar_ipp” dataset to “Sex.” Using ggplot and the new dataset we created through this, the numbers of males and females incarcerated within each state are shown and distinguished through color coding in the bar graph (where red represents female and blue represents male). Through the visualization provided by this bar graph, we can see the total number of people incarcerated in each state, as well as how many females and males make up this total. 


Bar Chart 3A: Total Population Incarcerated in Each State by Sex
Goal: To show sex differences in total incarcerated population in each state

```{r figure 3A, echo = T, message=F, warning=F}

#Pivot longer
state_data <- states_incar_ipp %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "Sex",
               values_to = "Value")
 
#plot              
state_total_stacked <- ggplot(state_data, aes(x = NAME, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Sex))+
  theme_linedraw()+
  labs(x = "State", y = "Total Population Incarcerated", 
       title = "Total Population Incarcerated in Each State by Sex")+
  coord_flip() 
  
#Show visualization
state_total_stacked

```

For Barchart 3B, we created a stacked bar chart to show the number of females incarcerated per every 100,000 people in the state, alongside the number of males incarcerated per every 100,000 people in the state. First, the variables in our state data were “pivoted longer” through the R function “pivot_longer” and the forward pipe operator to make the columns “Male Incarceration Ratio Per 100,000” and “Female Incarceration Ratio Per 100,000” from the “states_incar_ipp” dataset to “Sex.” Using ggplot and the new dataset created by this, the number of males incarcerated per 100,000 people in each state and the number of females incarcerated per 100,000 people within each state are shown and distinguished through color coding in the bar graph. In the graph, red represents “Female Incarceration Ratio Per 100,000” and blue represents “Male Incarceration Ratio Per 100,000”.  Through the visualization provided by this stacked bar graph, we can see the total number of each sex incarcerated per 100,000 people in each state by gender. Note, using case_match(), we renamed the variables to a shorter form, i.e. # of XX/100000, in order to save space on the graph. 

Bar Chart 3B: Number of People Incarcerated Per 100,000 People in the State's Population by Sex
Goal: To show the Number of females incarcerated per 100,000 people in each state by sex. This is important because this data takes into consideration total population size.  

```{r figure 3B, echo = T, message=F, warning=F}

#Pivot longer
gendered_states_incar_fmpc <- states_incar_ipp %>%
  pivot_longer(cols = c(`Number of Males Incarcerated Per 100,000 People`, 
                        `Number of Females Incarcerated Per 100,000 People`),
               names_to = "Sex",
               values_to = "Value") %>% mutate(Sex=
                case_match(Sex, "Number of Males Incarcerated Per 100,000 People" 
                           ~ "# of Male/100,000", 
                      "Number of Females Incarcerated Per 100,000 People" 
                      ~ "# of Female/100,000"))
 
#Plot              
state_total_stacked <- ggplot(gendered_states_incar_fmpc, aes(x = NAME, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Sex))+
  theme_linedraw()+
  labs(x = "State", y = "Number of People Incarcerated Per 100,000 People in the State's Population", 
       title = "Number of People Incarcerated/100,000 People in the State's 
Population by Sex")+
  coord_flip() +
  theme(legend.position = "bottom")
  
#Show visualization
state_total_stacked

```


### *Stacked Bar Charts 4A and 4B*

For Barchart 4A, we created a stacked bar chart to display the number of females incarcerated alongside the number of males incarcerated in each region. First, the variables in our state data were “pivoted longer” through the R function “pivot_longer” and the forward pipe operator to make the columns “Male” and “Female” from the cleaned dataset to “Sex.” Using ggplot and our new dataset “states_incar_region3”, the total numbers of males and females incarcerated within each state are shown and distinguished through color coding in the bar graph (where red represents female and blue represents male). Through the visualization provided by this bar graph, we can see the total number of people incarcerated in each region, as well as how many females and males compose these totals. 

Bar Chart 4A: Total Population Incarcerated in Each Region by Sex
Goal: To show sex differences in total incarcerated population in each region

```{r figure 4A, echo = T, message=F, warning= F}

#Pivot longer
states_incar_region3 <- states_incar_region2 %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "Sex",
               values_to = "Value")

#Plot 
sex_by_region <- ggplot(states_incar_region3, aes(x = Region, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Sex))+
  theme_linedraw()+
  labs(x = "Region", y= "Total Population Incarcerated in Each Region", 
       title = "Total Population Incarcerated in Each Region by Sex")+
  coord_flip() 
 
#Print plot  
sex_by_region

```

For Barchart 4B, we created a stacked bar chart to show the number of females incarcerated per every 100,000 people in the region, alongside the number of males incarcerated per every 100,000 people in the region. First, the variables in our region data were “pivoted longer” through the R function “pivot_longer” and the forward pipe operator to make the columns “Male Incarceration Ratio Per 100,000” and “Female Incarceration Ratio Per 100,000” from the BJS dataset to “Sex.” Using ggplot and the new dataset that we made in this process, the number of males and females incarcerated within each region per 100,000 people in the state are shown and distinguished through color coding in the bar graph. In the graph, red represents “Number of Females Incarcerated Per 100,000 People in Each Region” and blue represents “Number of Males Incarcerated Per 100,000 People in Each Region.” Note, using case_match(), on this graph as well, we renamed the variables to a shorter form i.e. # of XX/100000 inorder to save space on the graph.  Through the visualization provided by this bar graph, we can see the total number of people of the female and male sexes incarcerated per 100,000 people in each state. 


Bar Chart 4B: Total Number of People Incarcerated Per 100,000 People in Each Region by Sex
Goal: To show the Number of females incarcerated per 100,000 people in each region by sex. This is important because this data takes into consideration total population size.  

```{r figure 4B, echo=T, message=FALSE, warning=FALSE}

#Pivot longer
region_data_fpc <- states_incar_region2 %>%
  pivot_longer(cols = c(`Number of Females Incarcerated Per 100,000 People`, 
                        `Number of Males Incarcerated Per 100,000 People`),
               names_to = "Sex",
               values_to = "Value") %>% mutate(Sex=
                case_match(Sex, "Number of Males Incarcerated Per 100,000 People" 
                           ~ "# of Male/100,000", 
                      "Number of Females Incarcerated Per 100,000 People" 
                      ~ "# of Female/100,000"))

#Plot 
sex_by_region_pc_graph <- ggplot(region_data_fpc, aes(x = Region, y = Value))+
  geom_bar(position = "stack", stat = "identity",
           aes(fill = Sex))+
  theme_linedraw()+
  labs(x = "Region", y= "Number of People Incarcerated/100,000 people", 
       title = "Total Number of People Incarcerated/100,000 People in Each Region by Sex")+
  coord_flip() +
  theme(legend.position = "bottom")
 
#See Visualization  
sex_by_region_pc_graph
 
```

Having stacked bar charts by sex in both states and regions allows us to directly see how many males and females are composing each location’s total incarceration numbers. The data shows that overall, males are being incarcerated in higher numbers than females, but females are still being incarcerated at different rates across different regions and states. 

### *Spatial Graphs 1, 2, 3*

Utilizing the dataset set up earlier called “states_incar_ipp” with the variable “Female Incarceration Ratio Per 100,000”, we created a spatial map that displayed the number of females incarcerated per 100,000 people in each state. We use the color option “magma” in the viridis package where yellow represents the smaller ratios and a deep purple represents the highest ratios. We also filtered the coordinates using xlim so that Hawaii would be on the left side of the graph as opposed to the far right side of the scale. 

Spatial Graph 1: Number of Females Incarcerated per 100,000 People in Each State
Goal: To show the state data in new way using spatial analysis in hopes of finding trends based on location. 

```{r figure 5, echo = T, message=F, warning=F}

#Plot spatial and Number of Females Incarcerated Per 100,000 People in Each State data
ggplot(states_incar_fpc)+
  geom_sf(aes(fill = `Number of Females Incarcerated Per 100,000 People`))+
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(title = "Number of Females Incarcerated Per 100,000 People in Each State") +
  coord_sf(xlim = c(-185, -60)) +
  theme(legend.position = "bottom")

```

We created a new dataset called “states_incar_ftm” in which we divided the total number of females incarcerated in each state by the number of males incarcerated in each state using the preprocessed dataset “states_incar_ipp”. We then created a spatial map that displayed the female to male incarceration ratio of people in each state using the color option “magma” in the viridis package—yellow represents the smaller ratios (less females to males) and a deep purple represents the highest ratios (more females to males). The spatial data is from tidycensus and was added with “left_join()” earlier. We also filtered the coordinates using xlim so that Hawaii would be on the left side of the graph as opposed to the far right side for the scale. 

Spatial Graph 2: Number of Females Incarcerated per 100,000 people in each state to Number of Males Incarcerated per 100,000 People in Each State
Goal: To visualize comparisons between male and female incarceration rates spatially

```{r figure 6, echo = T, message=F, warning=F}

#Add Number of Females Incarcerated per 100,000 in State Population to 
#Number of Males Incarcerated per 100,000 in State Population

states_incar_ftm <- states_incar_ipp %>%
  mutate("Female per 100,000 population to Male per 100,000" = 
           `Number of Females Incarcerated Per 100,000 People`/ 
           `Number of Males Incarcerated Per 100,000 People`)

head(states_incar_ftm) #New data set with # of females incarcerated 
#in a state /# of males incarcerated in a state

#plot combined data - Female to Male
ggplot(states_incar_ftm)+
  geom_sf(aes(fill = `Female per 100,000 population to Male per 100,000`))+
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(title = "Number of Females Incarcerated/100,000 people per
Number of Males Incarcerated/100,000 people")  +
  coord_sf(xlim = c(-185, -60)) +
  theme(legend.position = "bottom")

```

We created a new dataset with only the states that have escalated/sometimes escalate prostitution to a felony called “strict_prostitution_states”. The states included were Oklahoma, Texas, Utah, Pennsylvania, Nevada, Kentucky, West Virginia, Vermont, Louisiana, Idaho, Indiana, Illinois. We classified these states as stricter states. In order to make this dataset, we utilized “states_incar_ipp”. We then used geom_sf() to layer the outlines of these states onto Spatial Graph 1 in order to create borders of these specific states. We edited the borders of these specific states to have a thicker line width, as well as made them green, so that they would stand out. This graph highlights the number of females incarcerated per 100,000 people in the state alongside the specific states with stricter prostitution stances to allow for comparisons between states' legal punishment of prostitution and female incarceration numbers. At a quick glance, there does not appear to be a correlation between the two. 

Spatial Graph 3: Highlighting states with specific policies on spatial graph of total # of female incarcerated to per 100,000 in each state
Goal: To provide a way to spatially analyze female state incarceration data alongside visual markers on states with stricter stances. 

```{r figure 7, echo = T, message=F, warning=F}

#Highlight specific states 

strict_prostitution_states <- states_incar_ipp %>% 
  filter(NAME %in% c("Oklahoma", "Texas", "Utah", "Pennsylvania", "Nevada", "Kentucky", 
                     "West Virginia", "Vermont", "Louisiana", "Idaho", "Indiana", 
                     "Illinois"))

head(strict_prostitution_states) #new data set for states with stricter prostitution laws

#Number of Females Incarcerated Per 100,000 People in Each State, Highlighting States 
#with Strict Prostitution Laws in Green

ggplot(states_incar_ipp)+
  geom_sf(aes(fill = `Number of Females Incarcerated Per 100,000 People`))+
  scale_fill_viridis(option = "magma", direction = -1)+
  labs(title = "Number of Females Incarcerated Per 100,000 People in Each State", 
       subtitle = "States with Stricter Punishment on Prostitution are Highlighted in 
Green Borders")+
  geom_sf(data = strict_prostitution_states, 
          fill = NA, 
          color = "green",
          size = 0.5) +
  coord_sf(xlim = c(-185, -60)) +
  theme(legend.position = "bottom")

```

### *T-Test*

Although the spatial map showed that there does not seem to be a correlation between state policies and female incarceration numbers, we wanted to verify this through another method. The last function we ran through R was intended to tie together our policy research on prostitution laws to the data we had extracted from BJS and the U.S. Census data. Research on state prostitution laws showed us that the states with the strictest prostitution laws were Oklahoma (recently increased the level of penalties on prostitution), as well as Texas, Utah, Pennsylvania, Nevada, Kentucky, West Virginia, Vermont, Louisiana, Idaho, Indiana, Illinois (as these states can turn prostitution offenses to felony charges under specific circumstances). We created a column in which each of these states were labeled as strict or not strict based on the classifications used for the above map. We then began the t-test calculation. In order to do this we used an independent two sample t-tests, which is a measure used to determine if there is a significant difference between the means of two different groups (in which the p-value is less than 0.05). We utilized the default t-test in R, which is a two-tailed Welch’s (unpaired) two-sample t-test. Given that the p-value was 0.1561 and it must be under 0.05 to show statistical significance, we conclude that in our test, strict prostitution laws were not statistically significant in explaining incarceration rates. 

T-Test: Highlighting states with specific policies on spatial graph of total # of female incarcerated to male incarcerated per state. If the p-value is less than 0.05 (or 5%), the result is considered statistically significant. 
Goal: See if there is a statistically significant difference in female incarceration tp population ratios between strict versus non-strict states. 

```{r t-test, echo = T, message=F, warning=F}

#Add row for states with stricter punishments on prostitution vs not
strictness_states_incar_fpi <- states_incar_ipp %>% 
  mutate(Strictness=case_match(NAME,
                             #Strict
                             c("Oklahoma", "Texas", "Utah", "Pennsylvania", "Nevada", 
                               "Kentucky", "West Virginia", "Vermont", "Louisiana", 
                               "Idaho", "Indiana", "Illinois") ~ "Strict",
                             c("South Carolina", "Maine", "New Hampshire", "Massachusetts", 
                               "Connecticut", "Rhode Island", "New York", "California", 
                               "New Jersey", "North Dakota", "South Dakota", "Nebraska", 
                               "Kansas", "Minnesota", "Iowa", "Missouri",  
                               "Wisconsin", "Michigan", "Ohio", "Washington", "Oregon", 
                               "Wyoming", "Montana", "Colorado", "Arizona", "New Mexico", 
                               "Arkansas", "Mississippi", "Tennessee", "Alabama", "Georgia", 
                               "Florida", "North Carolina", "Maryland", "Delaware", 
                               "Pennsylvania", "Alaska", "Hawaii", "Virginia") ~ "Not Strict")) %>%
  group_by(Strictness) %>% mutate(sum = sum(Total, na.rm = TRUE))

head(strictness_states_incar_fpi) #shows new column that organizes states as strict vs not strict

#T-test of strict vs not strict population Number of Females Incarcerated per 100,000 in 
#State Population
t.test(`Number of Females Incarcerated Per 100,000 People` ~ Strictness, 
       data = strictness_states_incar_fpi)


```

## Table of Contents

* <a href="index.html">Index</a>
* <a href="Introduction.html">Introduction</a>
* <a href="DataSources.html">Data Sources</a> 
* <a href="DataProcessingandMethods.html">Data Processing and Methods</a>
* <a href="Limitations.html">Limitations</a> 
* <a href="Conclusion.html">Conclusion</a> 
* <a href="ContributionsReferencesAppendix.html">Contributions, References, and Appendix</a> 
* <a href="CompleteReport.html">Complete Report</a> 